---
title: "Lab 1 - Introduction to R"
author: " "
date: " "
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: inline
utput:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
  html_notebook:
    toc: yes
    toc_depth: 3
    toc_float: yes
---

## Instructions and Overview

If you've reached this point in the lab, congratulations! You have successfully cloned your GitHub repo for this class into RStudio. At the end of the assignment, I will explain how to commit your changes and push them back to GitHub, where they will be accessible to me for grading.

As we move through this notebook, we will begin to get acquainted with R Markdown - the format in which these notebooks are written -, the themes of the course, and the statistical programming language R. Collectively, the eight labs in this course will give you an opportunity to engage with the course themes, data manipulation, and data presentation directly in R notebooks. Each week, we will take pieces of our labs to build upon this dashboard. 

For those of you that are Stats majors, some of the content of the labs will feel like review. Please keep in mind though that the primary goal of this course is not to teach you statistical analysis or data science. Instead, the primary goal of this course is to encourage you to think critically about a data practice - recognizing the human forces that are always involved in shaping data, responsibly documenting uncertainties in data, and communicating ways that data can both produce and delimit insight. While analyzing data in appropriate ways are important parts of each assignment, the reflections that you will write about the analyses you produce are perhaps even more important. 

## Introduction to RStudio

In this course we will be learning how to code in R inside the RStudio interface. You may be wondering the difference between the two. 

**R** is a statistical computing language that allows you to:
* Store data in a variety of formats
* Perform calculations on data and variables, 
* Build functions and applications, and 
* Transform and graphically represent data

**RStudio** is an Integrated Development Environment (IDE) for statistical computing. It is a platform for using and coding in R.

Before we continue, please watch this 8-minute video introducing you to the RStudio Environment. 

https://www.youtube.com/embed/TJmNvfhLCoI

### Installing Packages

The R community has developed a series of *packages* or collections of functions that we can use when coding in R. In this course we will be leveraging a series of packages known as the Tidyverse - designed to help users clean, transform, analyze, and visualize data. We will also be leveraging a series of packages for building a shiny data dashboard.

You will need to install the following packages in your R environment. One way of doing it, would be to copy and paste the following line into your Console. If you don't know where you console is, be sure to rewatch the RStudio YouTube video above. 

install.packages(c("tidyverse", "lubridate", "shiny", "shinydashboard", "shinyWidgets", "rmarkdown"))

### Loading Packages

Once packages are installed, to load a package, we call the function *library()*, which let's R know that we will be referencing the functions encoded for this package in our code. You can think of an R package like a book, stored in a computer library. When you call, library(), it is as if we are checking that book out of the library so that we may reference its functions in our code. Run the code below after installing the packages above to ensure that everything installed properly. Refer to the video above if you do not know how to run the code. 

```{r}
library(tidyverse)
library(lubridate)
library(shiny)
library(shinydashboard)
library(shinyWidgets)
```


## Introduction to R Markdown

If you are viewing this document in RStudio, you'll notice a series of characters (like hashtags and asterisks) being used to demarcate headers, code chunks, italicized and emphasized words, bullets, numbering, figures, and tables. These characters are encoding how the document should be presented when we publish it. These symbols are part of R Markdown - an authoring framework for R that enables us to style documents and embed code. When we publish this document, you see the text following or in between these symbols formatted in a standardized way. If you would like to see how this document will appear once it's published, click the 'Preview' button in the navigation bar above. This will preview the published document in the Viewer to the right. 

In future assignments, I have clearly marked where to fill in responses, so, for the most part, you will not necessarily need to change any of the R Markdown. I would, however, like to go over how R Markdown works so that you have a sense of how content will the notebook will appear in the published document. 

### Code Chunks

```{r eval=FALSE}
This grey box between ```{r} and ``` is a code chunk. 

Typically, when we put code in this chunk and then publish our R Markdown document, 
the code written here will be run and output on the screen. 

However, by placing eval=FALSE in the brackets, we tell R not to evaluate the code. 

This is useful when we are just filling text into the code chunk.
```

### Headings

```{r eval=FALSE}
Use '#' to create headings:
  
# Heading 1
## Heading 2
### Heading 3
```

### Tables

```{r eval=FALSE}
Tables can be formated like this:
  
Header 1 | Header 1
-------- | --------
Cell 1  | Cell 2
Cell 3  | Cell 4
```

Header 1 | Header 1
-------- | --------
Cell 1  | Cell 2
Cell 3  | Cell 4

### Bullets and Numbering

```{r eval=FALSE}
You can bullet content using the '*':

* Point 1
* Point 2
```

This will output:

* Point 1
* Point 2

```{r eval=FALSE}
You can number content like this:

1. Point 1
2. Point 2
```

This will output:

1. Point 1
2. Point 2


For more information about formatting documents in R Markdown see [this](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf) cheatsheet. Note how I created a hyperlink here!

## Introduction to R

Now we will begin to review some basic functionality in R. As you read through each section, be sure to run the code chunks to see how the code is operating. 

### Assigning Variables

Variables are used to store data in R. We use "<-" to assign a variable. The text that comes before "<-" will be our variable name, and the text that comes after "<-" will be the value stored in it. 
```{r}
class_name <- "Data Sense and Exploration"
class_dept <- "STS"
class_number <- "115"
class_size <- 11

cs_majors <- 3
stats_majors <- 2
other <- 6
```

Every object in R has a particular *class*, which designates the variable's "type" and how functions can be applied to it. We can check the class of a variable by calling it in "class()".
```{r}
class(class_name)
class(class_size)
class(class_number)
```
Did you notice how class_number above is a character variable? This is because when we created the class_number variable above, we put the number in quotation marks - indicating to R to treat the number as a set of characters rather than a number. Since our class number is a reference to our class, it acts more like a label than a number. 

If someone drops a class, we can easily change the value of the variable.
```{r}
#Store the calculation in a variable.
class_size <- class_size - 1

#Print the variable
class_size
```
Notice how we re-assigned the new subtracted value to class_size above. 

We can also perform calculations on variables, such as addition, subtraction, multiplication, and division. Checking whether variables are greater than, less than, or equal to each other will return TRUE or FALSE. 
```{r}
cs_majors + stats_majors

cs_majors > stats_majors
```

We can also perform operations on strings - concatenating them with the *paste* function. For instance, we can paste the class_dept string with the class_number string to create a class code. When we do this, we need to tell R what characters should separate the strings. We will review this again later in the quarter. 
```{r}
class_code <- paste(class_dept, class_number, sep=" ")
class_code
```

### Vectors

A vector is a set of values (organized like a list) that are all of the same type. A vector can be of type integer, double, characters, logical, for example. We create a vector by placing a set of values in "c(___)". Here c stands for the function "combine" - indicating that we are combining values.  
```{r}
birth_months <- c(4, 7, 12, 3, 1)
first_letter_name <- c("A","B","C","D","E")
time_on_phone <- c(34, 90, 2, 6, NA)
```

We can measure the length of a vector by calling the function *length*. This counts how many values are listed in the vector. 
```{r}
length(birth_months)
```

We can also extract values from specific positions in the vector by referencing the index in brackets. So let's say I want to extract the third value in the vector. 
```{r}
birth_months[3]
```

To extract all values except the value in a particular position, we will use the "-" sign before the index in brackets. So let's say I want to extract all values except the third in the vector.
```{r}
birth_months[-3]
```

We can also extract a range of values from specific positions in the vector by referencing that range of indexes in brackets separated by a ":". So let's say I want to extract the first through the third value in the vector.
```{r}
first_letter_name[1:3]
```

Finally, to extract values from a non-sequential combination of specific positions in the vector, we can reference each of their indexes in brackets in "c()".  
```{r}
first_letter_name[c(1,3,5)]
```

We can perform operations on vectors - finding their max, their min, their sum, their average, for example. However, we may get an error if we have any empty values in the dataset. To avoid this, we need to communicate to R to remove NA values. 
```{r}
max(time_on_phone)
min(time_on_phone)
sum(time_on_phone)
mean(time_on_phone)

max(time_on_phone, na.rm=TRUE)
min(time_on_phone, na.rm=TRUE)
sum(time_on_phone, na.rm=TRUE)
mean(time_on_phone, na.rm=TRUE)
```

### Matrices

We won't be using matrices much in this course. However, knowing this data type will help you understand its differences from the data types we will use. Let's create a second numeric vector. 
```{r}
birth_months2 <- c(11, 12, 3, 4, 4)
```

By binding this together with birth_months, we create a matrix - or a 2-dimensional collection of elements of all the same type.
```{r}
birth_months_matrix <- rbind(birth_months, birth_months2)
birth_months_matrix
```

We can determine the number of rows in the matrix by calling *nrow* and we can determine the number of columns in the matrix by calling *ncol*.
```{r}
nrow(birth_months_matrix)
ncol(birth_months_matrix)
```

We can also extract values from specific positions in the vector by referencing the row position and the column position of the number in brackets. The format for doing so is [row, column] So let's say that we want to extract the value in the second row, third column.
```{r}
birth_months_matrix[2,3]
```

To extract the entire second row, we would leave the column position blank. This would return a *vector* of values stored in the second row of the matrix.
```{r}
birth_months_matrix[2,]
```

And to extract the entire third column, we would leave the row position blank. This would return a *vector* of values stored in the third column of the matrix. 
```{r}
birth_months_matrix[,3]
```

### Lists

I will only briefly go over *lists* because, for the most part, we will not be using them in this class. *Lists* are collections of objects in R. For instance, you can have a collection of numeric vectors, character vectors, logical vectors, matrices, and other lists. You can assign names to the objects in lists so that you can more easily reference them. For instance, below I assign the name x, y, and z to the three objects respectively. Once a name has been assigned to the object, you can reference it by listing the name of the list followed by the *$* followed by the name of the object. 

```{r}
first_list <- list(x = first_letter_name, y = time_on_phone, z = birth_months)

first_list$x[2:3]
first_list$y[c(1,4)]
first_list$z[2]
```

### Data Frames

In this course, we will be working primarily with rectangular datasets in a data type called a *data frame*.  A data frame has a certain number of rows and a certain number of columns. It is a list of *vectors* of all the same length. 

**From this point forward, we will refer to each row in a data frame as an *observation* and each column in a data frame as a *variable*.** This is because rows refer to something that we see in the world, and columns describe that thing we are seeing. Imagine we have a table like this below.

Name | Age | Birth Month | Time on Phone
---- | ---- | ---- | ---- |
Sally | 23 | 3 | 42
Julie | 40 | 2 | 98
Mark | 14 | 8 | 120

Each row is an observation - in this case a person - and each column is a variable describing something about a person. Note how each column is also a vector. The Name column is a character vector of names. The Age column is a numeric vector of ages.

Let's go head and create this vector below:

```{r}
df <- data.frame(name = c("Sally", "Julie", "Mike"), 
                 age = c(23, 40, 14), 
                 birth_month = c(3, 2, 8),
                 time_on_phone = c(42, 98, 120))
```

When working with very large datasets, we need tools to help us get a sense of the dataset without having to load the entire data frame. For instance, we can view the first 6 rows of the dataset by calling *head()*.
```{r}
head(df)
```

*dim()* will tell us the dimensions of the data frame - i.e. the number of rows and the number of columns in the data frame. 
```{r}
dim(df)
```

"str()" provides a great deal of information about the observations in the data frame, including the number of variables, the number of observations, the variable names, their data types, and a list of observations.
```{r}
str(df)
```

Just like as we had done with matrices, we can extract particular rows and columns in a data frame by referring to their indexes in brackets. 
```{r}
df[2,4]
```

We often don't want to have the count the index of each column in order to refer to a particular variable in our dataframe. Instead, we can refer to the variable (column) name using the same "$" notation that we discussed above. For instance, I could see the values in the birth_month column by calling:
```{r}
df$birth_month
```

If I wanted the extract the second observation in the birth_month, I would call:
```{r}
df$birth_month[2]
```

What if you don't know what the column names are? To see a list of column names, we could refer to the data dictionary. We could also use the function "colnames()". 
```{r}
colnames(df)
```
Finally, on the top right hand side of your RStudio screen, you will see a tab called Environment. If you click on that tab, you should see all of the data that you have stored in variables. You can check column names here. 



##Practice (5 pts.)

Before you submit this assignment, you should name it. Let's briefly practice some of the tricks that you have learned. First, create three variables: one that contains your name, another one with your email, and a last one with your age. You can do so by filling in the blanks ("_____") below. These variables should be of different classes, two of them will be "character" and one "numeric" (Hint: you can use the class() function to check the class of a variable). Finally, print the three variables.

```{r}

name <- ________
________ <- ________
________ <- ________

#What function helps you determining their classes?
________ 

#Now, print your three variables:
________
________
________

```
**BONUS (0.3 pts)**

You are almost done! The last thing that I want you to do is to transform this notebook into a sharable document, which can be easily uploaded and opened on Canvas. If you complete this question, you will receive an extra 0.3 points. 

To do so, we are going to use the rmarkdown package to transform your R notebook into a PDF file. 
First, click the “Knit” icon that appears on the Rstudio UI above your file in the scripts editor. 
A drop down menu will let you select the type of output that you want. For this assignment, please select "Knit to PDF" (but you could also transform your R notebook file into an HTML, PDF, or Word document).
When you click the button, Rstudio (via the) will duplicate your text in the new file format. rmarkdown will use the formatting instructions that you provided with markdown syntax (in the case of this lab, automatically generated) Once the file is rendered, RStudio will show you a preview of the new output and save the output file in your working directory.

If this is not the case, it may be that there is an error in the code. To correct such error, please fill the blanks in the former question (name, email, age). If you want to learn more about knitting, I suggest watching this video: https://www.youtube.com/watch?v=GZKJDp8r7Uc


## Submitting your dcoument on Canvas

Congratulations! With this, we conclude this week's lab. Now, it is time to upload your documents and submit them through Canvas. First, make sure you saved your R notebook (cntrl + S / or 'File' > 'Save'). Now go to our STS 115 Canvas site. Then, go to the "Assignments" tab. There you will see "Lab #1" (5 pts). Upload your RMD file. You can 'add another file.' If you knitted and exported your notebook as a PDF, please upload the PDF file to receive your extra bonus (0.3 pts).
